Supplement to ``{\myTitle}''

### 2024-02-03 si/si.Rnw
updated to match arxiv submission

@@

R0_M6 <- function (params, be = T) 
{
    a = params[ifelse(be, "alpha_be1", "alpha_af1")]
    b = params[ifelse(be, "Beta_be1", "Beta_af1")]
    D = params["D1"]
    mu = params[ifelse(be, "mu_be1", "mu_af1")]
    unname((a + (1 - a) * mu) * b * D)
}

plot_si <- function(x, U=373, limit=4, order="population") {
  df <- as.data.frame(x)
  df[x@unit_obsnames] <- log10(df[x@unit_obsnames]+1)
  # change the order, sort by population
  if (order=="population"){
    popu <- df$pop[1:U]
    index <- order(order(popu, decreasing = T))
  } else if (order=="cases"){
    index <- 1:U
  }
  wuhan_index <- which(df$city=="Wuhan")
  df$Wuhan <- NA
  df$Wuhan[wuhan_index] <- df$cases[wuhan_index]
#  shenyang_index <- which(df$city=="Shenyang")
#  df$Shenyang <- NA
#  df$Shenyang[shenyang_index] <- df$cases[shenyang_index]
  df$y_ticks <- rep(index,30)

  ggplot2::ggplot(data = df,
    mapping = ggplot2::aes(x = !!rlang::sym(x@timename), y = .data$y_ticks)) +
    ggplot2::scale_x_continuous(expand=c(0,0)) +
    ggplot2::scale_y_continuous(expand=c(0,0), sec.axis = sec_axis(~ .)) +
    ggplot2::geom_tile(mapping = ggplot2::aes(fill = !!rlang::sym(x@unit_obsnames))) +
    ggplot2::scale_fill_gradient(low = "#FFFFFF", high = "#000000",limits=c(0,limit)) +
    ggplot2::labs(x = "time",
                  y = "unit",
                  fill = paste("log10\n(",x@unit_obsnames,"+1)",sep="")) +
    ggnewscale::new_scale("fill") +
    ggplot2::geom_tile(mapping = ggplot2::aes(fill = .data$Wuhan), data = ~subset(., !is.na(Wuhan))) +
    ggplot2::scale_fill_gradient(low = "#FFFFFF", high = "#FF0000",limits=c(0,limit)) + 
    ggplot2::labs(fill = "log10\n(Wuhan+1)") +
    ggnewscale::new_scale("fill") +
    ggplot2::theme(
      axis.text.y=ggplot2::element_blank(),
      axis.ticks.y=ggplot2::element_blank(),
      panel.border=ggplot2::element_blank(),
      axis.line.y = ggplot2::element_line(linewidth = 0.5, linetype = "solid"))

@@

%% for flow diagrams
\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary {arrows.meta}
\usetikzlibrary{shapes.geometric}

\newcommand\LiMobility{M$_1$}
\newcommand\LiParams{M$_2$}
\newcommand\BenchmarkIID{M$_3$}
\newcommand\BenchmarkAR{M$_4$}
\newcommand\RevisedModelUnconstrained{M$_5$}
\newcommand\RevisedModelConstrained{M$_6$}
\newcommand\MainResultsTable{1}

%% for algorithm pseudocode
\usepackage{enumerate,alltt,xstring}
\usepackage[ruled,noline,linesnumbered]{algorithm2e}
\SetKwFor{For}{for}{do}{end}
\SetKwFor{While}{while}{do}{end}
\SetKwInput{KwIn}{input}
\SetKwInput{KwOut}{output}
\SetKwInput{KwCplx}{complexity}
\SetKwInput{KwIndices}{note}
\SetKwBlock{Begin}{procedure}{}
\DontPrintSemicolon
\newcommand\argequals{{\,=\,}}
\newcommand\data[1]{#1^*}
\newcommand\giventh{{\hspace{0.5mm};\hspace{0.5mm}}}
\newcommand\seq[2]{{#1}\!:\!{#2}}
\newcommand\mydot{{\,\cdot\,}}
\newcommand\given{{\,\vert\,}}
\newcommand\normal{{\mathrm{Normal}}}
\newcommand\vmeasure{V}
\newcommand\emeasure{e}

\newcommand\Time{N}
\newcommand\ttime{n}
\newcommand\Np{J}
\newcommand\np{j}
\newcommand\altNp{q}

\newcommand\prob{\mathrm{Prob}}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[version=4]{mhchem}
\usepackage{stmaryrd}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{float}
\usepackage{subfigure}
\usepackage{hyperref}
\usepackage{fullpage}
\usepackage{xurl}
\usepackage{makecell}
\usepackage{multirow}
\usepackage{multicol}
\usepackage[ruled,noline,linesnumbered]{algorithm2e}
\usepackage[normalem]{ulem}% to use \sout in feedback commands
\usepackage{bm}
\usepackage[mathscr]{euscript}
\usepackage{amsthm}
\usepackage{natbib}
\usepackage{bbm} % for \mathbbm{1}

\newcommand\lik{L}
\newcommand\loglik{\ell}
\newcommand\myvec[1]{{\mathbf{#1}}}
\newcommand\proglang[1]{\textsf{#1}} 
\newcommand\code[1]{\texttt{#1}}
\newcommand\pkg[1]{\code{#1}}
\newcommand\class[1]{\code{#1}}
\newcommand\unit{u}
\newcommand\altUnit{\tilde{u}}
\newcommand\Unit{U}     
\newcommand\rate{\mu}
\usepackage{upgreek}
%\newcommand\relativeTransmission{\phi}
\newcommand\relativeTransmission{\upmu}
\newcommand\diagnosisDelay{T_d}
\newcommand\latency{Z}
\newcommand\infectiousPeriod{D}
\newcommand\reportRate{\alpha}
\newcommand\Rzero{\mathcal{R}_0}
\newcommand\mobilityFactor{\vartheta}
\newcommand\eqsep{\hspace{1mm}}
\newcommand\pop{P}
\newcommand\EzeroDescription{Initial $E$ in Wuhan}
\newcommand\AzeroDescription{Initial $A$ in Wuhan}
\newcommand\alphaDescription{Reported fraction}
\newcommand\before{\mathrm{be}}
\newcommand\after{\mathrm{af}}

\newcommand\E{\mathbb{E}}
\newcommand\Var{\mathrm{Var}}
\newcommand\var{\Var}

% for figures illustrating POMPs and the problem space for spatPomp
\usepackage{tikz,pgfplots}
\pgfplotsset{compat=1.16}
\usepgfplotslibrary{fillbetween}
\usetikzlibrary{patterns}
\usetikzlibrary{arrows,chains,matrix,positioning,scopes,fit,decorations.pathreplacing}
\tikzset{join/.code=\tikzset{after node path={%
\ifx\tikzchainprevious\pgfutil@empty\else(\tikzchainprevious)%
edge[every join]#1(\tikzchaincurrent)\fi}}}
%
% \tikzset{>=stealth,every on chain/.append style={join},
%         every join/.style={->}}
\tikzstyle{labeled}=[execute at begin node=$\scriptstyle,
   execute at end node=$]
\newcommand{\FixedLengthArrow}{1,0}

@@

\usepackage{color}
%orange for EI
\definecolor{orange}{rgb}{1,0.5,0}
\newcommand\ei[2]{\sout{#1} \textcolor{orange}{#2}}
\newcommand\eic[1]{\textcolor{orange}{[#1]}}
%green for PN
\definecolor{green}{rgb}{0,0.5,0}
\newcommand\pn[2]{\sout{#1} \textcolor{green}{#2}}
\newcommand\pnc[1]{\textcolor{red}{[#1]}}
%purple for JF
\definecolor{purple}{rgb}{0.5,0,1}
\newcommand\jf[2]{\sout{#1} \textcolor{purple}{#2}}
\newcommand\jfc[1]{\textcolor{purple}{[#1]}}
%teal for TODO
\definecolor{teal}{rgb}{0,0.5,0.8}
\newcommand\todo[1]{\textcolor{teal}{[TO DO? #1]}}

@@

\begin{document}
%\SweaveOpts{concordance=TRUE}

\date{\small Compiled \today,  using \proglang{R} \Sexpr{paste0(R.version$major,".",R.version$minor)}, \pkg{metapoppkg} \Sexpr{packageVersion('metapoppkg')}, \pkg{spatPomp} \Sexpr{packageVersion('spatPomp')}, and \pkg{pomp} \Sexpr{packageVersion('pomp')}.}
\title{Supplement to ``Machine Learning for Mechanistic Models of Metapopulation Dynamics''}
\author{Jifan Li, Edward L. Ionides, Aaron A. King, Mercedes Pascual and Ning Ning}
\maketitle

@@

if (!dir.exists("../model1/si")) {
  dir.create("../model1/si", recursive = TRUE)
}
@

@@

R0_uc_be <- readRDS("../profile9/R0/profile_R0_be_full.rds")[,c("logLik","R0_be1")]
R0_uc_be <- R0_uc_be[R0_uc_be$R0_be1<22.5,]
# rbind(
#   readRDS("../profile9/R0/out_3/profile_R0_be.rds")[,c("logLik","R0_be1")],
#   readRDS("../profile9/R0/out_3/profile_R0_be2.rds")[,c("logLik","R0_be1")]
# )

R0_uc_af <- readRDS("../profile9/R0/profile_R0_af_full.rds")[,c("logLik","R0_af1")]
# R0_uc_af <- rbind(
#   readRDS("../profile9/R0/out_3/profile_R0_af.rds")[,c("logLik","R0_af1")],
 #   readRDS("../profile9/R0/out_3/profile_R0_af2.rds")[,c("logLik","R0_af1")]
 # )

Z_uc_be <-  readRDS("../profile9/Z/profile_Z_be_full.rds")[,c("logLik","Z_be1")]
# rbind(
#   readRDS("../profile9/Z/out_3/profile_Z_be3.rds")[,c("logLik","Z_be1")]
# )

Z_uc_af <-   readRDS("../profile9/Z/profile_Z_af_full.rds")[,c("logLik","Z_af1")]
# rbind(
#   readRDS("../profile9/Z/profile_Z_af_full.rds")[,c("logLik","Z_af1")]
# )

D_uc_be <-  readRDS("../profile9/D/profile_D_be_full.rds")[,c("logLik","D_be1")]
# rbind(
#   readRDS("../profile9/D/profile_D_be_full.rds")[,c("logLik","D_be1")]
# )

D_uc_af <-   readRDS("../profile9/D/profile_D_af_full.rds")[,c("logLik","D_af1")]
D_uc_af <- D_uc_af[D_uc_af$D_af1>1.1,]
# rbind(
#   readRDS("../profile9/D/profile_D_af_full.rds")[,c("logLik","D_af1")]
# )

alpha_uc_be <- readRDS("../profile9/alpha/profile_alpha_be_full.rds")[,c("logLik","alpha_be1")]
# rbind(
#   readRDS("../profile9/alpha/out_3/profile_alpha_be2.rds")[,c("logLik","alpha_be1")],
#   readRDS("../profile9/alpha/out_3/profile_alpha_be3.rds")[,c("logLik","alpha_be1")]
# )
# alpha_uc_be <- alpha_uc_be[alpha_uc_be$alpha_be1>0.02 & alpha_uc_be$alpha_be1<0.2 ,]

alpha_uc_af <-   readRDS("../profile9/alpha/profile_alpha_af_full.rds")[,c("logLik","alpha_af1")]
# rbind(
#   readRDS("../profile9/alpha/out_3/profile_alpha_af2.rds")[,c("logLik","alpha_af1")]
# )
# alpha_uc_af <- alpha_uc_af[alpha_uc_af$alpha_af1>0.2 ,]

@@

R0_c_be <- rbind(
  readRDS("../profile7/R0/out_3/profile_R0_be.rds")[,c("logLik","R0_be1")],
  readRDS("../profile7/R0/out_3/profile_R0_be2.rds")[,c("logLik","R0_be1")]
)
# R0_c_be <- R0_c_be[R0_c_be$R0_be1>2 & R0_c_be$R0_be1<6  ,]

@@

D_c <- rbind(
  readRDS("../profile7/D/out_3/profile_D.rds")[,c("logLik","D1")],
  readRDS("../profile7/D/out_3/profile_D2.rds")[,c("logLik","D1")] 
)
# D_c <- D_c[D_c$D1>2.5 & D_c$D1<8  ,]
#myplot_c(D_c)

@@

Beta_c_af <- rbind(
  readRDS("../profile7/Beta/out_3/profile_Beta_af.rds")[,c("logLik","Beta_af1")],
  readRDS("../profile7/Beta/out_3/profile_Beta_af2.rds")[,c("logLik","Beta_af1")]
)
# Beta_c_af <- Beta_c_af[Beta_c_af$Beta_af1>0.1 ,]
#myplot_c(Beta_c_af)

@@

We give the mathematical description of the COVID-19 models described in Figure~1 and Table~1 of the main text.
These metapopulation models are partially observed Markov process (POMP) models with additional spatial structure, known as SpatPOMP models \citep{asfaw23arxiv}.
The diagram in Figure~\ref{fig:spatpomp} defines a general SpatPOMP corresponding to a metapopulation with $U$ spatial units.
The latent dynamic state is $\myvec{X}(t)=\big(X_1(t),\dots,X_U(t)\big)$, which we also write as $\myvec{X}(t)=X_{1:U}(t)$.
The observation model at time $t_n$ is $\myvec{Y}_{n}=\big(Y_{1,n},\dots,Y_{U,n}\big)=Y_{1:U,n}$, for $n$ taking values in $\seq{1}{N}$.
The data, $\myvec{y}^*_n=y^*_{1:U,n}$, are modeled as a realization of the random variable $Y_{1:U,n}$.

@@

Each transition $dN_{VW}$ has an associated rate, $\mu^{}_{VW}$, which may depend on the state of other compartments, or on covariate processes, on parameters, or on time.
For all compartments other than the source/sink compartment, $T$, it is convenient to specify rates per capita.
For transitions which enter a compartment from $T$, we specify a total rate.
The non-zero transition rates are therefore as follows:
\begin{eqnarray}
\label{eq:muSE}
\rate_{S_{\unit}E_{\unit}} &=& \beta S_{\unit}(t) \left(\frac{I_{\unit}(t)+ \relativeTransmission A_{\unit}(t)}{\pop_{\unit}(t)}\right)
d\Gamma_{\unit}/dt,
\\
\rate_{S_{\unit}T} \eqsep = \eqsep \rate_{E_{\unit}T} \eqsep = \eqsep \rate_{A_{\unit}T} &=& \mobilityFactor \sum_j \frac{M_{\unit j}(t)}{\pop_{\unit}-I_{\unit}},
\\
\rate_{TS_{\unit}} &=& \mobilityFactor \sum_j \frac{M_{j \unit}(t)\, S_{j}}{\pop_{j}-I_{j}},
\\
\rate_{TE_{\unit}} &=& \mobilityFactor \sum_j \frac{M_{j \unit}(t)\, E_{j}}{\pop_{j}-I_{j}},
\\
\rate_{TA_{\unit}} &=&\mobilityFactor \sum_j \frac{M_{j \unit}(t)\,  A_{j}}{\pop_{j}-I_{j}},
\\
\rate_{E_{\unit}A_{\unit}} \eqsep = \eqsep \rate_{E_{\unit}I_{\unit}}  &=&  \alpha/Z,
\\
\rate_{A_{\unit}R_{\unit}} \eqsep = \eqsep \rate_{I_{\unit}R_{\unit}}  &=& 1/D,
\\
\rate_{C^a_{\unit}C^b_{\unit}} \eqsep = \eqsep \rate_{C^b_{\unit}C_{\unit}}  &=& 2\diagnosisDelay.
\label{eq:muCbC}
\end{eqnarray}
We define $d\Gamma_{\unit}/dt$ in (\ref{eq:muSE}) as non-negative multiplicative gamma white noise with variance parameter $\sigma_{SE,\unit}$.
That is, $\Gamma_{\unit}(t)$ is a gamma process with stationary independent increments, such that $\Gamma_{\unit}(t)-\Gamma_{\unit}(s)$ is gamma distributed with mean $t-s$ and variance $\sigma_{SE}(t-s)$.
Equations (\ref{eq:dS})-(\ref{eq:muCbC}) therefore specify an overdispersed continuous time Markov process via the limit of a discrete time Euler approximation as the discretization step approaches zero \citep{breto09,breto11}.

@@

Our models {\LiMobility}, {\LiParams}, {\RevisedModelUnconstrained} and {\RevisedModelConstrained} are extensions of the model of \citet{li20}.
The original model of \citet{li20}, which we call \code{li20}, represents the dynamic model by a system of ordinary differential equations with random rates, as discussed further in Section~\ref{sec:li20}.
The general model including {\LiMobility}, {\LiParams}, {\RevisedModelUnconstrained} and {\RevisedModelConstrained}, which we call \code{li23}, represents the models as continuous-time Markov chains.
In addition to this change,  \code{li23} considers two model aspects not investigated by \citet{li20}: (i) overdispersed process noise; (ii) an adjusted movement matrix to ensure that all cities are connected.
In  {\LiMobility}, these two features are turned off, and so this model is very similar to \code{li20}, as documented subsequently in Section \ref{sec:li20} (and specifically Figure~\ref{fig:li20_plot}) where we discuss \code{li20} in more detail. {\LiMobility} and {\LiParams} have the constraint that $\sigma_{SE}=0$.
{\LiMobility} has an additional constraint that $\tau=0$ since this parameter was not included in the dynamic model of \cite{li20}.
Another difference between these models is that {\LiMobility} uses the mobility data from \citep{li20} whereas {\LiParams} (together with {\RevisedModelUnconstrained} and {\RevisedModelConstrained}) use the modification described in Section~\ref{sec:mobility}.
{\RevisedModelUnconstrained} involves estimation of all the parameters estimated by \cite{li20}, with the addition of $\sigma_{SE}$ and $\tau$.
{\RevisedModelConstrained} adds the additional constraints that $\latency^{\after}=\latency^{\before}$  and $\infectiousPeriod^{\after}=\infectiousPeriod^{\before}$.

@@

  \Sexpr{myround(if(PROF_MAX) attr(R0_be_plot_c,"mle") else R0_M6(mle6,be=T),2)} &

@@

  \Sexpr{myround(if(PROF_MAX) attr(R0_af_plot_c,"mle") else R0_M6(mle6,be=F),2)} &

@@

<<panel_plot, message=FALSE, echo=FALSE,fig.width=8.5,fig.height=10,out.width="6.5in",fig.cap="Daily case report time series for 373 cities: (A) the real data; (B) a simulation from model {\\LiMobility}; (C) a simulation from model~{\\RevisedModelConstrained}. Within each panel, cities are ordered by population, largest on the bottom row.">>=
i <- 3
library(metapoppkg)
plotdir <- paste0("plot_",i,"/")
if (!dir.exists(plotdir)) dir.create(plotdir)
stew(file=paste0(plotdir,"panel.rda"),{
  U <-switch(i, 5,50,373)
  M5 <- li23(U=U)
  M5_sim <- simulate(M5,params=mle3,seed=100)
  M1 <- spatPomp(li23(U=U,version="li20period3",for_ibpf=F, mob_modify_factor=0),
    rmeasure=spatPomp_Csnippet(
      unit_statenames='C',
      code="
        double* cases=&cases1;
        int u;
        for (u = 0; u < U; u++) {
          if (C[u] > 0.0) {
            cases[u] = nearbyint(C[u]);
          } else {
            cases[u] = 0.0;
          }
        } 
      "
    )
  )
  coef(M1)["sigma_SE1"] <- 0
  M1_sim <- simulate(M1,seed=100)
})
panel_plot <- bake(file=paste0(plotdir,"panel2.rds"),{
  limit <- log10(max(c(obs(M5),obs(M5_sim),obs(M1_sim)))+1)
  M5_plot <- plot_si(M5_sim, U=U, limit=limit) + ggtitle(expression("(C) "~Model~M[5])) +
    ylab("City") +
    xlab("Day") +
    theme(legend.position = "bottom")
  M1_plot <- plot_si(M1_sim, U=U, limit=limit) + ggtitle(expression("(B) "~Model~M[1])) +
    ylab("City") +
    theme(legend.position = "none", axis.title.x=element_blank(),
      axis.ticks.x=element_blank(), axis.text.x=element_blank())
  data_plot <- plot_si(M5, U=U, limit=limit) + ggtitle(expression("(A) "~Data)) + 
    ylab("City") +
    theme(legend.position = "none", axis.title.x=element_blank(),
      axis.ticks.x=element_blank(), axis.text.x=element_blank())
  data_plot/M1_plot/M5_plot
})  
panel_plot
@

<<percentile_plot, message=FALSE, echo=FALSE,fig.width=8.5,fig.height=10,out.width="6.5in",fig.cap="Simulated daily case reports for model {\\RevisedModelUnconstrained}, showing the 10th, 50th and 90th percentiles. Within each panel, cities are ordered by population, largest on the bottom row.">>=
i <- 3
stew(file=paste0(plotdir,"percentile_df.rda"),{
  percentile_df <- sim_percentile_si(U=switch(i,5,50,373),Nsim = switch(i,10,100,100), params=mle3)
})  
percentile_plot <- plot_percentile_si(percentile_df)
percentile_plot
@

@@

Our estimate of the relative transmissibility of unreportable to reportable cases is considerably higher than \cite{li20}.
Indeed, our estimate during the first time period is $\relativeTransmission^{\before}=\Sexpr{myround(mle3["mu_be1"],2)}$.
The data are compared to simulations from models {\LiMobility} and {\RevisedModelUnconstrained} in Figure~\ref{fig:panel_plot}. 
The greater number of cases for model {\RevisedModelUnconstrained} compared to  {\LiMobility} is explained by its higher initial $\Rzero$.
The variability in {\RevisedModelUnconstrained} is explicitly included in the model and fitted to the data; it therefore matches the variability in the data more closely than  {\LiMobility}.
If many simulations are made from {\RevisedModelUnconstrained}, the pointwise 10$^{\mathrm{th}}$ percentile is similar to a simulation from {\LiMobility} (Figure~\ref{fig:percentile_plot}).
The similarity between model {\LiMobility} and \code{li20} is evident by comparing Figure~\ref{fig:panel_plot}(B) with Figure~\ref{fig:li20_plot} in Section~\ref{sec:li20}.
The code and parameters for the simulation from \code{li20} are taken directly from \citet{li20}.

@@

\clearpage

@@

stew(file=paste0(plotdir,"li20_plot.rda"),{
  i <- 3
  U <-switch(i, 5,50,373)
  M1 <- spatPomp(li23(U=U,version="li20period3",for_ibpf=F, mob_modify_factor=0),
               rmeasure=spatPomp_Csnippet(
                 unit_statenames='C',
                 code="
        double* cases=&cases1;
        int u;
        for (u = 0; u < U; u++) {
          if (C[u] > 0.0) {
            cases[u] = nearbyint(C[u]);
          } else {
            cases[u] = 0.0;
          }
        } 
      "
    )
  )
  coef(M1)["sigma_SE1"] <- 0
  M1_sim <- simulate(M1,seed=100)

@@

  lps_plot=lps_plot + ggtitle("LPS") + 
    ylab("City") +
   theme(legend.position = "none", axis.title.x=element_blank(),
        axis.ticks.x=element_blank(), axis.text.x=element_blank())
  M2_plot <- plot_si(M2_sim, U=U, limit=limit) + ggtitle("M2") +
    ylab("City") +
    xlab("Day") +
    theme(legend.position = "bottom")
  M1_plot <- plot_si(M1_sim, U=U, limit=limit) + ggtitle("M1") +
    ylab("City") +
    theme(legend.position = "none", axis.title.x=element_blank(),
        axis.ticks.x=element_blank(), axis.text.x=element_blank())
  rm("M1","M1_sim","M2","M2_sim")
})
M1_plot <- M1_plot + ggtitle(expression("(B) "~Model~M[1]))
M2_plot <- M2_plot + ggtitle(expression("(C) "~Model~M[2]))
lps_plot <- lps_plot + ggtitle("(A)  Model of Li et al. (2020)")
lps_panel <- lps_plot/M1_plot/M2_plot
lps_panel
@

@@

>>>>

\input{../header}
\usepackage{natbib}
\usepackage{fullpage}

@@

@@

@@

@@

\begin{document}

\date{\small Compiled \today,  using \proglang{R} \Sexpr{paste0(R.version$major,".",R.version$minor)}, \pkg{metapoppkg} \Sexpr{packageVersion('metapoppkg')}, \pkg{spatPomp} \Sexpr{packageVersion('spatPomp')}, and \pkg{pomp} \Sexpr{packageVersion('pomp')}.}
\title{Supplement to ``{\myTitle}''}
\author{\myAuthorList}
\maketitle

@@

if (!dir.exists("../model1/si")) {
  dir.create("../model1/si", recursive = TRUE)
}

i <- 3
plotdir <- paste0("plot_",i,"/")
if (!dir.exists(plotdir)) dir.create(plotdir)

@@

R0_uc_be <- readRDS("../profile9/R0/profile_R0_be_full.rds")[,c("logLik","R0_be1")]
R0_uc_be <- R0_uc_be[R0_uc_be$R0_be1<22.5,]
R0_uc_af <- readRDS("../profile9/R0/profile_R0_af_full.rds")[,c("logLik","R0_af1")]
Z_uc_be <-  readRDS("../profile9/Z/profile_Z_be_full.rds")[,c("logLik","Z_be1")]
Z_uc_af <-   readRDS("../profile9/Z/profile_Z_af_full.rds")[,c("logLik","Z_af1")]
D_uc_be <-  readRDS("../profile9/D/profile_D_be_full.rds")[,c("logLik","D_be1")]
D_uc_af <-   readRDS("../profile9/D/profile_D_af_full.rds")[,c("logLik","D_af1")]
D_uc_af <- D_uc_af[D_uc_af$D_af1>1.1,]
alpha_uc_be <- readRDS("../profile9/alpha/profile_alpha_be_full.rds")[,c("logLik","alpha_be1")]
alpha_uc_af <-   readRDS("../profile9/alpha/profile_alpha_af_full.rds")[,c("logLik","alpha_af1")]

@@

R0_c_be <- rbind(
  readRDS("../profile7/R0/out_3/profile_R0_be.rds")[,c("logLik","R0_be1")],
  readRDS("../profile7/R0/out_3/profile_R0_be2.rds")[,c("logLik","R0_be1")]
)

@@

D_c <- rbind(
  readRDS("../profile7/D/out_3/profile_D.rds")[,c("logLik","D1")],
  readRDS("../profile7/D/out_3/profile_D2.rds")[,c("logLik","D1")] 
)
# D_c <- D_c[D_c$D1>2.5 & D_c$D1<8  ,]

@@

Beta_c_af <- rbind(
  readRDS("../profile7/Beta/out_3/profile_Beta_af.rds")[,c("logLik","Beta_af1")],
  readRDS("../profile7/Beta/out_3/profile_Beta_af2.rds")[,c("logLik","Beta_af1")]
)
# Beta_c_af <- Beta_c_af[Beta_c_af$Beta_af1>0.1 ,]

@@

We give the mathematical description of the COVID-19 models described in Figure~{\MainFigModelDiagram} and Table~{\MainResultsTable} of the main text.
These metapopulation models are partially observed Markov process (POMP) models with additional spatial structure, known as SpatPOMP models \citep{asfaw23arxiv}.
The diagram in Figure~\ref{fig:spatpomp} defines a general SpatPOMP corresponding to a metapopulation with $U$ spatial units.
The latent dynamic state is $\myvec{X}(t)=\big(X_1(t),\dots,X_U(t)\big)$, which we also write as $\myvec{X}(t)=X_{1:U}(t)$.
The observation model at time $t_n$ is $\myvec{Y}_{n}=\big(Y_{1,n},\dots,Y_{U,n}\big)=Y_{1:U,n}$, for $n$ taking values in $\seq{1}{N}$.
The data, $\myvec{y}^*_n=y^*_{1:U,n}$, are modeled as a realization of the random variable $Y_{1:U,n}$.

@@

Each transition $dN_{VW}$ has an associated rate, $\mu^{}_{VW}$, which may depend on the state of other compartments, or on covariate processes, on parameters, or on time.
For all compartments other than the source/sink compartment, $T$, it is convenient to specify rates per capita.
For transitions which enter a compartment from $T$, we specify a total rate.
The non-zero transition rates are therefore as follows:
\begin{eqnarray}
\label{eq:muSE}
\rate_{S_{\unit}E_{\unit}} &=& \beta \left(\frac{I_{\unit}(t)+ \relativeTransmission A_{\unit}(t)}{\pop_{\unit}(t)}\right)
d\Gamma_{\unit}\big/dt,
\\
\rate_{S_{\unit}T} \eqsep = \eqsep \rate_{E_{\unit}T} \eqsep = \eqsep \rate_{A_{\unit}T} &=& \mobilityFactor \sum_j \frac{M_{\unit j}(t)}{\pop_{\unit}-I_{\unit}},
\\
\rate_{TS_{\unit}} &=& \mobilityFactor \sum_j \frac{M_{j \unit}(t)\, S_{j}}{\pop_{j}-I_{j}},
\\
\rate_{TE_{\unit}} &=& \mobilityFactor \sum_j \frac{M_{j \unit}(t)\, E_{j}}{\pop_{j}-I_{j}},
\\
\rate_{TA_{\unit}} &=&\mobilityFactor \sum_j \frac{M_{j \unit}(t)\,  A_{j}}{\pop_{j}-I_{j}},
\\
 \eqsep \rate_{E_{\unit}I_{\unit}}  &=&  \alpha/Z,
 \\
 \rate_{E_{\unit}A_{\unit}} \eqsep &=& (1-\alpha)\big/ Z
\\
\rate_{A_{\unit}R_{\unit}} \eqsep = \eqsep \rate_{I_{\unit}R_{\unit}}  &=& 1/D,
\\
\rate_{C^a_{\unit}C^b_{\unit}} \eqsep = \eqsep \rate_{C^b_{\unit}C_{\unit}}  &=& 2\big/\diagnosisDelay.
\label{eq:muCbC}
\end{eqnarray}
We define $d\Gamma_{\unit}/dt$ in (\ref{eq:muSE}) as non-negative multiplicative gamma white noise with variance parameter $\sigma_{SE,\unit}$.
That is, $\Gamma_{\unit}(t)$ is a gamma process with stationary independent increments, such that $\Gamma_{\unit}(t)-\Gamma_{\unit}(s)$ is gamma distributed with mean $t-s$ and variance $\sigma_{SE}(t-s)$.
Equations (\ref{eq:dS})-(\ref{eq:muCbC}) therefore specify an overdispersed continuous time Markov process via the limit of a discrete time Euler approximation as the discretization step approaches zero \citep{breto09,breto11}.

@@

Our models {\LiMobility}, {\LiParams}, {\RevisedModelUnconstrained} and {\RevisedModelConstrained} are extensions of the model of \citet{li20}.
The original model of \citet{li20}, which we call \code{li20}, represents the dynamic model by a system of ordinary differential equations with random rates, as discussed further in Section~\ref{sec:li20}.
The general model including {\LiMobility}, {\LiParams}, {\RevisedModelUnconstrained} and {\RevisedModelConstrained}, which we call \code{li23}, represents the models as continuous-time Markov chains.
In addition to this change,  \code{li23} considers two model aspects not investigated by \citet{li20}: (i) overdispersed process noise; (ii) an adjusted movement matrix to ensure that all cities are connected.
In  {\LiMobility}, these two features are turned off, and so this model is very similar to \code{li20}, as documented in Section \ref{sec:li20}, below, which discusses \code{li20} in more detail.
{\LiMobility} and {\LiParams} have the constraint that $\sigma_{SE}=0$.
{\LiMobility} has an additional constraint that $\tau=0$ since this parameter was not included in the dynamic model of \cite{li20}.
Another difference between these models is that {\LiMobility} uses the mobility data from \citep{li20} whereas {\LiParams} (together with {\RevisedModelUnconstrained} and {\RevisedModelConstrained}) use the modification described in Section~\ref{sec:mobility}.
{\RevisedModelUnconstrained} involves estimation of all the parameters estimated by \cite{li20}, with the addition of $\sigma_{SE}$ and $\tau$.
{\RevisedModelConstrained} adds the additional constraints that $\latency^{\after}=\latency^{\before}$  and $\infectiousPeriod^{\after}=\infectiousPeriod^{\before}$.

@@

  \Sexpr{myround(if(PROF_MAX) attr(R0_be_plot_c,"mle") else R0v3(mle6,be=T),2)} &

@@

  \Sexpr{myround(if(PROF_MAX) attr(R0_af_plot_c,"mle") else R0v3(mle6,be=F),2)} &

@@

<<percentile_plot, message=FALSE, echo=FALSE,fig.width=8.5,fig.height=10,out.width="6.5in",fig.cap="Simulated daily case reports for model {\\RevisedModelUnconstrained}, showing the 10th, 50th and 90th percentiles. Within each panel, cities are ordered by population, largest on the bottom row.">>=
stew(file=paste0(plotdir,"percentile_df.rda"),{
  percentile_df <- sim_percentile_si(U=switch(i,5,50,373),Nsim = switch(i,10,100,100), params=mle3)
})  
percentile_plot <- plot_percentile_si(percentile_df)
percentile_plot
@

@@

The data are compared to simulations from models {\LiMobility} and {\RevisedModelConstrained} in Figure~{\MainFigSimulations} of the main text.
The variability in {\RevisedModelUnconstrained} and {\RevisedModelConstrained} is explicitly included in the model and fitted to the data; it therefore matches the variability in the data more closely than  {\LiMobility}.
If many simulations are made from {\RevisedModelUnconstrained}, the pointwise 10$^{\mathrm{th}}$ percentile is similar to a simulation from {\LiMobility} (Figure~\ref{fig:percentile_plot}).
The similarity between model {\LiMobility} and \code{li20} is evident by comparing Figure~{\MainFigSimulations}(B) with Figure~\ref{fig:li20_plot} in Section~\ref{sec:li20}.
The code and parameters for the simulation from \code{li20} are taken directly from \citet{li20}.

@@

@@

stew(file=paste0(plotdir,"li20_plot.rda"),{
  U <-switch(i, 5,50,373)
  M1 <- spatPomp(li23(U=U,version="li20period3",for_ibpf=F, mob_modify_factor=0),
               rmeasure=spatPomp_Csnippet(
                 unit_statenames='C',
                 code="
        double* cases=&cases1;
        int u;
        for (u = 0; u < U; u++) {
          if (C[u] > 0.0) {
            cases[u] = nearbyint(C[u]);
          } else {
            cases[u] = 0.0;
          }
        } 
      "
    )
  )
  coef(M1)["sigma_SE1"] <- 0
  M1_sim <- simulate(M1,seed=100)

@@

  lps_plot=lps_plot + ggtitle("LPS") + 
    ylab("City") +
   theme(legend.position = "none", axis.title.x=element_blank(),
        axis.ticks.x=element_blank(), axis.text.x=element_blank())
  M2_plot <- plot_li(M2_sim, U=U, limit=limit) + ggtitle("M2") +
    ylab("City") +
    xlab("Day") +
    theme(legend.position = "bottom")
  M1_plot <- plot_li(M1_sim, U=U, limit=limit) + ggtitle("M1") +
    ylab("City") +
    theme(legend.position = "none", axis.title.x=element_blank(),
        axis.ticks.x=element_blank(), axis.text.x=element_blank())
  rm("M1","M1_sim","M2","M2_sim")
})
M1_plot <- M1_plot + ggtitle(expression("(B) "~Model~M[1]))
M2_plot <- M2_plot + ggtitle(expression("(C) "~Model~M[2]))
lps_plot <- lps_plot + ggtitle("(A)  Model of Li et al. (2020)")
lps_panel <- lps_plot/M1_plot/M2_plot
lps_panel
@

@@

%%%%%%%%%%%% Review of inference methods %%%%%%%%%%%%%%

\section{\secSep Review of inference methods for metapopulation models}
\label{sec:review}
